NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-02-26 main.tex
Update text and pdfs

The {\it application layer} is composed of repositories and tools that read information from the {\it databank layer} for further analyses, and their results. Because the tools in the {\it application layer} do not interfere with the {\it databank layer}, they can be freely extended by anyone for wide range of purposes. Here we exemplify this with two For example, the analyses described in below sections locate in an application layer repository at \url{www.github.com/NMRLipids/DataBankManuscript}, and content of the databank can be accessed with a graphical user interface tool at~\url{www.databank.nmrlipids.fi/}. A more detailed description on the NMRlipids databank structure is available the supplementary information.

@@

\subsection{NMRlipids databank-GUI: graphical access to the MD simulation data}\label{section:access}
NMRlipids databank-GUI provides easy access to the content of the NMRlipids databank via graphical user interface (GUI) that is available at~\url{www.databank.nmrlipids.fi/}. The content of the NMRlipids databank can be easily browsed and adanced searched options based on systems content and properties are available. Systems and their basic properties can be visualized and compared between different simulations and experiments. Simulations can be also ranked simulations based on their quality evaluated against experimental data.

\subsection{NMRlipids databank-API: programmatic access to the MD simulation data}
The NMRlipids databank-API, available at \url{www.github.com/NMRLipids/Databank}, enables automatic analysis of virtually any property from all simulations in the NMRlipids databank. The NMRlipids databank-API delivers functions that enable straightforward implementation of the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}A. Iteration through the README.yaml files in the {\it databank layer} enables the access to the raw data of each simulation within the databank via the links stored in these files. The affiliation of universal molecule and atom names to the simulation specific names are delivered for each simulation in README.yaml and mapping files, which enables automatic analyses over all systems independently on the specific naming convention used in a simulation. Finally, the analysis results are stored using the same structure as in the {\it databank layer}.

While analysis codes and results for basic membrane properties 
%area per lipid, C–H bond order parameters, X-ray scattering form factors and thickness 
are included in the {\it Databank layer}, unlimited amount of further analyses can be implemented by anyone in separate repositories in the {\it application layer}. {\it Application layer} repositories that are organized using the {\it Databank layer} structure can be straightforwardly accessed programmatically and fuhther analyzed by implementing the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}B by using the NMRlipids databank-API functions. An {\it application layer} repository used to produce the novel analyses from the NMRlipids databank exemplified in the next section is available at~\url{www.github.com/NMRLipids/DataBankManuscript}.

@@

\subsection{Applications of the NMRlipids databank}
\subsubsection{Selection of simulation parameters using the NMRlipids databank: Best models for most abundant neutral membrane lipids}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
Understanding of how closely simulations resemble the reality is crucial for selecting best models for specific applications, assessing the reliability and significance of MD simulation results, and using atomistic simulations in parametrization of coarse grained models. However, undefined universal quality measures for MD simulations and the lack of a prevailing force field that would perform best in all simulations has led to many controversial results in membrane simulations~\cite{antila22b}. 
%For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. 
For example, Fig.~\ref{fig:POPCPOPEapls}C illustrates divergent predictions from different force fields for the lateral packing of POPC and POPE membranes that are the two most biologically abundant neutral membrane lipids~\cite{vanmeer08}. We have previously resolved such controversies for membrane structures and ion binding affinities for neutral~\cite{botan15,catte16} and charged lipids~\cite{antila19,bacle21} by comparing simulations with C-H bond order parameters from NMR, yet this requires manual comparisons between large number of simulations. To streamline this process,
%the selection of best simulations for specific applications, and to guide the development of atomistic and coarse grained parameters,
we have defined quantitative quality measures, based on C--H bond order parameters from NMR and X-ray scattering form factors, for lipid bilayer simulations in the NMRlipids databank. 
%The quality for each simulation in the NMRlipids databank with the corresponding experimental data available is currently done using C--H bond order parameters from NMR experiments and form factors from X-ray scattering experiments. 
The C--H bond order parameters are sensitive to conformational ensembles of individual lipid molecules~\cite{ollila16}, while correlating well also with the membrane thickness and lateral packing as shown in Fig.~\ref{fig:quality}F. The X-ray scattering form factors are related to membrane dimensions \textit{via} the electron density profile. In particular, locations of minima in these form factors are correlated with the membrane lateral packing and thickness as shown in Figs.~\ref{fig:quality}F,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}. For quantitative quality measures, probabilities for each order parameter to locate within experimental error, and the distance of the first form factor minimum from the experimental value are calculated from a simulation.
%For detailed definitions of quality measures, see the supplementary information.
Together with the API, these enable rapid and automatic quality evaluation of simulations in the NMRlipids databank, thereby promoting the selection of best simulations for specific applications without laborious force field evaluation, and automatic parametrization procedures for atomistic and coarse grained simulations.

To select the best models for POPC and POPE among simulations compared in Fig.~\ref{fig:POPCPOPEapls}C, we show the ranking based on C--H bond order parameter quality in Fig.~\ref{fig:quality}A for the best performing simulations in the NMRlipids databank, and for the best performing POPE lipid bilayer simulations in Fig.~\ref{fig:quality}B. Direct comparisons with experiments are shown for selected simulations in Figs.~\ref{fig:quality}C--E and~\ref{fig:POPC_POPE_dataSI}. 
%The power of the NMRlipids databank in selecting the best simulation parameters for a specific application is demonstrated for mixtures of PE and PC lipids in Fig.~\ref{fig:POPCPOPEapls} showing area per lipids from available POPC and POPE bilayer simulations in the databank from different force fields at 310\,K. Predictions from different force fields deviate significantly in terms of absolute values and the slope of decrease upon addition of POPE. 
Because the area per lipid correlates with the average order parameter of the $\textit{sn}$-1 chain (Fig.~\ref{fig:quality} F), quality evaluation in Fig.~\ref{fig:quality} and direct comparison in Fig.~\ref{fig:POPC_POPE_dataSI} can be used to select the simulations that best capture area per lipids for POPE and POPC lipids in Fig.~\ref{fig:POPCPOPEapls}C. In these results, simulations with Slipids parameters ranks 1\textsuperscript{st} for POPC with a clear difference to others, and 2\textsuperscript{nd} for POPE with only marginally lower quality than GROMOS-CKP. Simulations with CHARMM36 and GROMOS-CKP parameters predict overly packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while the area per lipid for POPC is overestimated in Lipid17. For POPE, CHARMM36 and Lipid17 predict membranes that are too packed. In conclusion, the quality evalution based on the NMRlipids databank suggests that Slipids parameters are the best available choice for simulations with PC and PE lipids, at least for the cases were membrane packing is relevant.

>>>>

The {\it application layer} is composed of repositories and tools that read information from the {\it databank layer} for further analyses. Because {\it application layer} does not interfere with the {\it databank layer}, it can be freely extended by anyone for wide range of purposes. Here we demonstrate this with two examples: NMRlipids databank graphical user interface (NMRlipids databank-GUI) tool at~\url{www.databank.nmrlipids.fi/} and a repository exemplifying novel analyses utilizing NMRlipids databank as discussed below (\url{www.github.com/NMRLipids/DataBankManuscript}). More detailed description of the NMRlipids databank structure is available in the supplementary information.

@@

\subsection{NMRlipids databank-GUI: graphical access to the MD simulation data}\label{section:access}
NMRlipids databank-GUI provides easy access to the content of the NMRlipids databank via graphical user interface (GUI) that is available at~\url{www.databank.nmrlipids.fi/}. The NMRlipids databank-GUI has advanced searching options for system content and properties. The search results are ranked based on their quality evaluation against experimental data which enables rapid selection of best available simulations for specific systems.  Systems can be also visualized and their properties can be compared with different simulations and experiments.

\subsection{NMRlipids databank-API: programmatic access to the MD simulation data}
The NMRlipids databank-API, available at \url{www.github.com/NMRLipids/Databank}, enables automatic analysis of virtually any property from all simulations in the NMRlipids databank. This enables wide range of novel analyses that take the advantage on unprecendented amount MD simulation data with programmatic access. This is realized in the NMRlipids databank-API by delivering functions that enable straightforward implementation of the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}A. Iteration through the README.yaml files in the {\it databank layer} enables the access to the raw data of each simulation within the databank via the links stored in these files. The affiliation of universal molecule and atom names to the simulation specific names are delivered for each simulation in README.yaml and mapping files, which enables automatic analyses over all systems independently on the specific naming convention used in a simulation. Finally, the analysis results are stored using the same structure as in the {\it databank layer}.

While analysis codes and results for basic membrane properties 
%area per lipid, C–H bond order parameters, X-ray scattering form factors and thickness 
are included in the {\it Databank layer}, unlimited amount of further analyses can be implemented by anyone in separate repositories in the {\it application layer}. {\it Application layer} repositories that are organized using the {\it Databank layer} structure can be straightforwardly accessed programmatically and further analyzed by implementing the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}B by using the NMRlipids databank-API functions. Novel analyses demonstrating the power of the NMRlipids databank in selecting best simulation models, analyses of rare phenomena and extending MD simulations to new fields are implemented in an {\it application layer} repository located at~\url{www.github.com/NMRLipids/DataBankManuscript}.

@@

\subsection{Applications of the NMRlipids databank}
\subsubsection{Selection of simulation parameters using the NMRlipids databank: Best models for most abundant neutral membrane lipids}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
%
%Understanding of how closely simulations resemble the reality is crucial for selecting best models for specific applications, assessing the reliability and significance of MD simulation results, and using atomistic simulations in parametrization of coarse grained models. However, undefined universal quality measures for MD simulations and the lack of a prevailing force field that would perform best in all simulations has led to many controversial results in membrane simulations~\cite{antila22b}. 
%
%For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. 
%
%Understanding of how closely simulations resemble the reality is crucial for selecting best models for specific applications,
The quality of lipid bilayer MD simulations has to be carefully assessed in their applications to minize the detrimental consequences of artificial MD simulation results for lipid bilayers~\cite{antila22b}. This is possible, for example, by using C--H bond order parameters from NMR and X-ray scattering form factors \cite{ollila16}, although it requires comparisons between large number of simulations which is laborious even with collaborative approaches~\cite{botan15,catte16,antila19,bacle21}. The quantitative quality measures in the NMRlipids databank substantially streamline this process by enabling automatic ranking of lipid bilayer simulations based on their quality against experiments. Here we demonstrate this by selecting the best models for simulations of membranes with POPC and POPE lipids that are the two most biologically abundant neutral membrane lipids~\cite{vanmeer08}.

Divergent predictions for the lateral packing of POPC and POPE membranes from different force fields is illustrated in Fig.~\ref{fig:POPCPOPEapls}C. To select the best models for these membranes, the ranking based on C--H bond order parameter quality for the best performing POPC and POPE simulations in the NMRlipids databank are shown in Figs.~\ref{fig:quality}A and B, respectively. Direct comparisons with the experimental are shown in Figs.~\ref{fig:quality}C--E and~\ref{fig:POPC_POPE_dataSI} for the most relevant simulations. Based on these rankings, simulations with Slipids parameters are the most realistic for POPC lipids with a clear difference to others, and 2\textsuperscript{nd} most realistic for POPE with only marginally lower quality than GROMOS-CKP simulations. Simulations with CHARMM36 and GROMOS-CKP parameters predict overly packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while the area per lipid for POPC is overestimated in Lipid17. For POPE, CHARMM36 and Lipid17 predict membranes that are too packed. 
Because the area per lipid correlates with the average order parameter of the $\textit{sn}$-1 chain (Fig.~\ref{fig:quality} F), quality evaluation in Fig.~\ref{fig:quality} and direct comparison in Fig.~\ref{fig:POPC_POPE_dataSI} can be used to select the simulations that best capture area per lipids for POPE and POPC lipids in Fig.~\ref{fig:POPCPOPEapls}C.
In conclusion, the quality evalution based on the NMRlipids databank suggests that Slipids parameters are the best available choice for simulations with PC and PE lipids, at least for the cases were membrane packing is relevant.

The C--H bond order parameters are sensitive to conformational ensembles of individual lipid molecules~\cite{ollila16}, while correlating well also with the membrane thickness and lateral packing as shown in Fig.~\ref{fig:quality}F. The X-ray scattering form factors are related to membrane dimensions \textit{via} the electron density profile. In particular, locations of minima in these form factors are correlated with the membrane lateral packing and thickness as shown in Figs.~\ref{fig:quality}F,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}. For quantitative quality measures, probabilities for each order parameter to locate within experimental error, and the distance of the first form factor minimum from the experimental value are calculated from a simulation.
%For detailed definitions of quality measures, see the supplementary information.
Together with the API, these enable rapid and automatic quality evaluation of simulations in the NMRlipids databank, thereby promoting the selection of best simulations for specific applications without laborious force field evaluation, and automatic parametrization procedures for atomistic and coarse grained simulations.
